---
title: "Working With Dates"
author: "David Asare Kumi"
date: "9/30/2019"
output: html_document
---


## Starting Simple

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(1234)
d1 <-date()
d1
class(d1)
# Date Class
d2 <-Sys.Date()
d2
class(d2)

```

## Formatting Dates

- %d = day as a number (0-31).

- Weekday

1. %a = abbreviated weekday.

2. %A = unabbreviated weekday.

- Month

1. %m = month(00-12).

2. %b = abbreviated month.

3. %B = unabbreviated month.

- Year

1. %y = two(2) digit year.

2. %Y = four(4) digit year.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(1234)
format(d2,"%a %b %d")
format(d2,"%A %B %d")

```

### Creating Dates

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-c("1jan1960","2jan1960","31mar1960","30jul1960")
z <-as.Date(x,"%d%b%Y")
z
#Time difference of -1 day
z[1]-z[2]
as.numeric(z[1]-z[2])

```

### Converting to Julian

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
weekdays(d2)
months(d2)
julian(d2)

```

- The number of days that have occured since the origin day (1970) i.e. 1970-01-01.

### Lubridate

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(lubridate)
ymd("20140108")
mdy("08/04/2013")
dmy("03-04-2013")

```

- Lubridate convert a number to date. Lubridate is very important because you don't have to format the date.

### Dealing with Times

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
ymd_hms("2011-08-03 10:15:03")
ymd_hms("2011-08-03 10:15:03",tz="Pacific/Auckland")

```

- ?Sys.timezone gives you more information on time zones.

### Some functions have slightly different Syntax

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-dmy(c("1jan2013","2jan2013","31mar2013","30jul2013"))
wday(x[1])
wday(x[1],label=TRUE)

```

- lubridate; wday.

### Notes and Further Resources

- More information in this nice lubridate tutorial http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/

- The lubridate vignette is the same content http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html

- Ultimately you want your dates and times as class "Date" or the classes "POSIXct","POSIXlt". For more information type ?POSIXlt.

### Data Resources

### Open Government Sites

- United Nations http://data.un.org/

- U.S. http://www.data.gov/ -list of cities/states with open data

- United Kingdom http://data.gov.uk/

- France http://www.data.gouv.fr/

- Ghana http://data.gov.gh/

- Australia http://data.gov.au/

- Germany https://www.govdata.de/

- Hong Kong http://www.gov.hk/en/theme/psi/datasets/

- Japan http://www.data.go.jp/

- Many more http://www.data.gov/opendatasites

- Gapminder also has a large number of datasets available http://www.gapminder.org/

- Survey data from the United States http:www.asdfree.com/ This site gives you information about how to access the various data.

- Infochimps Market place http://www.infochimps.com/marketplace

- Kaggle (They offer datascience competitions) http://www.kaggle.com/

### Collections by data Scientists

- Hilary Mason http://bitly.com/bundles/hmason/1

- Peter Skomoroch https://delicious.com/psomoroch/dataset

- Jeff Hammerbacher http://www.quora.com/Jeff-Hammerbacher/Introduction-to-Data-Science-Data-Sets

- Gregory Piatetsky-Shapiro http://www.kdnuggets.com/gps.html

- http://blog.mortardata.com/post/67652898761/6-dataset-lists-curated-by-data-scientists

### More Specialized Collections

- Stanford Large Network Data.

- UCI Machine Learning.

- KDD Nugets Datasets.

- CMU Statlib.

- Gene expression omnibus.

- ArXiv Data.

- Public Datasets on Amazon Web Services.

### Some API's with R interfaces

- twitter and twitter R package.

- figshare and rfigshare.

- PLoS and rplos.

- rOpenSci.

- Facebook and RFacebook.

- Google maps and RGoogleMaps.

NB: You can use gsub to remove commas. gsub(pattern=",",replacement="",x=GDP).

## Exporting data from R

- For excel, you will need the xlsReadWrite package.

1. To a tab delimited text file;

- write.table(mydata,"C:/mydata.txt",sep="\t")

2. To an Excel spreadsheet;

- library(xlsx)

- write.xlsx(mydata,"C:/mydata.xlsx")

### To SPSS

- library(foreign)

- write.foreign(mydata,"C:/mydata.txt","C:/mydata.spss",package="SPSS")

### To SAS

- library(foreign)

- write.foreign(mydata,"C:/mydata.txt","C:/mydata.sas",package="SAS")

### To Stata

- library(foreign)

- write.dta(mydata,"C:/mydata.dta")

### To Word

- Create a table or data frame in R.

- Write this table to a comma separated .txt file using write.table().

- Copy and paste the content of the .txt file into word.

- In word,

1. Select the text you just pasted from the .txt file.

2. Go to table, convert, convert text to table.

3. Make sure "commas" is selected under "Separate text at".

4. Click OK.

### Example

- write.table(mtcars,file="mtcars.txt",sep="\t",row.names=TRUE,colnames=NA).

- Copy and paste and change the font to a monoSpace font like;

1. consolas.

2. Lucida console.

3. write.csv(mtcars,file="mtcars.csv").

4. write.table(mtcars,file="mtcars.txt",sep=",",quote=FALSE,row.names=F).

5. write.table(Data to Export,file="Data to Export.csv",sep=",",row.names=FALSE).

6. Double click on the file and it will open in excel by default. Set row.names=FALSE to get rid of the row names.

7. write.csv(mydata,file="mydata.csv"). Here exclude sep=",".


## Import Data from SPSS

- Package need to be installed.

- library(foreign).

- df <- read.spss("dataset.sav", use.value.label=TRUE, to.data.frame=TRUE).

- df is the name of data frame created in R, and dataset.sav is the file name of SPSS dataset we want to import, and use.value.label=TRUE to convert variables with value labels in SPSS into R factors, and to.data.frame=TRUE to make as data frame.

## Import Data from Stata

- To import a dataset from Stata into R, the function read.dta() from foreign package is used.

- library(foreign).

- df <- read.dta("dataset.dta").

- df is the name of data frame in R, and dataset.dta is the file name of Stata dataset we want to import.

## Import Data from SAS

- To import a dataset from SAS into R there are different methods, but most recommended is to export first the dataset from SAS into CSV and then to import in R.

- First use the code below in SAS (not R) to export data:

- run in SAS

- proc export data=dataset

- outfile="datast.csv"

- dbms=csv;

- run;

- Now that your data is exported you can import in R by using the code below:

- df <- read.csv("dataset.csv",header=T,as.is=T).

- Another way to upload SAS files (.XPT) direct to R environment is by using Hmisc package.

- library(Hmisc)

- df <- sasxport.get("/filename.xpt")

## Import Data from CSV

- df <- read.table("dataset.csv", header=TRUE, sep=",").

- df <- read.csv("dataset.csv",header=T,as.is=T).

## Import data from TXT

- df <- read.table("dataset.txt", as.is=TRUE, header=T).

- df <- read.table("dataset.txt", header=TRUE, sep=",").

- df <- read.table("dataset.txt", header=T, strings=F).

## Load Rdata in R

- save(df, file="mydata.Rdata").

- load("mydata.Rdata").

- To load Rdata in R is easy and straightforward method. However, first we need to know how to save the dataframe in R. The function used for saving the dataframe is save(objectlist, file="myfile"), where objectlist is the name of your current dataframe and myfile is the filename of RDATA you will save on your computer. The function to upload the Rdata in R is load().

- Other interesting functions are ls() which list the objects in the current workspace, rm(objectlist) which delete objects in your workspace.
