---
title: "Plotting Systems in R"
author: "David Asare Kumi"
date: "June 7, 2019"
output: html_document
---


We have three plotting systems in R

- Base Plotting System

- ggplot2

- Lattice

##The Base Plotting System

- "Artist's pellete" model

- Start with blank canvas, and build up from there

- Start with plot function (or similar)

- Use annotation functions to add/modify (text,lines,points,axis)

###Key notes

- The base plotting system is the oldest plotting system

- There are functions that generate the plots and other functions annotate the plots

###Advantages

- Convenient.

- Mirrors how we think of building plots and analyzing data.

###Drawbacks

- Cannot go back once plot has started (i.e. to adjust margins);need to plan in advance.

- Difficult to translate to others once a new plot has been created (no graphical language).

- Plot is just a series of R commands.

##Base Plot
```{r,echo=TRUE,message=FALSE,warning=FALSE}
#Plot a scatterplot
library(datasets)
data(cars)
dim(cars)
names(cars)
summary(cars)
str(cars)
with(cars,plot(speed,dist,cex=2,pch=20,col="red",xlab="Speed",ylab="Dist",
               main="Scatterplot of Dist on Speed"))
#Plot multiple boxplots
data(InsectSprays)
names(InsectSprays)
dim(InsectSprays)
str(InsectSprays)
table(InsectSprays$spray)
boxplot(count~spray,data=InsectSprays,col="yellow",xlab="Spray",ylab="Count",
        main="Boxplot of Count on Spray")

```

##The Lattice Plotting System

This is implemented in the lattice package

- Plots are created with a single function call (xyplot,bwplot,etc).

- Most useful for conditioning types of plots: Looking at how y changes with x across levels of z.

- Things like margins,spacing are set automatically because entire plot is specified at once.

- Good for putting many plots on a screen.

###Key notes

- Here you have to specify a lot of information.

- Lattice plots are panel plots.

###Downside/Drawbacks

- Sometimes awkward to specify an entire plot in a single function call.

- Annotation in plot is not especially intuitive.

- Use of panel functions and subscripts difficult to wield(use,employ) and requires intense preparation.

- Cannot add to plot once it is created.

##Lattice Plot

```{r,echo=TRUE,message=FALSE,warning=FALSE,fig.width=10}
library(lattice)
data(USMortality)
dim(USMortality)
names(USMortality)
summary(USMortality)
xyplot(Rate~Sex|Cause,data=USMortality,layout=c(5,2))

```

##ggplot2 System
This comes from a grammar of graphics (language for plotting)

- Splits the difference between base and lattice in a number of ways.

- Automatically deals with spacings, text, titles, but allows you to annotate by "adding" to a plot.

- Superficial similarity to lattice but generally easier and more intuitive to use.

- Default mode makes many choices for you (but you can still customize to your heart's desire).

- This is implemented in R in the ggplot2 package.

- The ggplot2 is also useful for conditioning plots.

##ggplot2 Plot

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
data(mpg)
dim(mpg)
names(mpg)
glimpse(mpg)
table(mpg$manufacturer)
table(mpg$model)
table(mpg$cyl)
table(mpg$fl)
table(mpg$year)
#Plot graphs 
qplot(displ,hwy,data=mpg,main="Scatterplot of Displ and Hwy")
qplot(displ,hwy,data=mpg,main="Scatterplot of Displ and Hwy",facets=~cyl)
qplot(displ,hwy,data=mpg,main="Scatterplot of Displ and Hwy",facets=~fl)

```


