---
title: "Control Structures in R"
author: "David Asare Kumi"
date: "10/24/2019"
output: html_document
---


## Introduction

- Control structures in R allow you to control the flow of execution of the program, depending on run time conditions.

- Common structures are;

1. If, else: testing a condition.

2. for: execute a loop a fixed number of times.

3. while: execute a loop while a condition is true.

4. repeat: execute a an infinite loop.

5. break: break the execution of a loop.

6. next: skip an interation of a loop.

7. return: exit a function.

- Most control structures are not used in interactive sessions, but rather when writing functions or longer expressions.

## Control Structures: if - else

- Control Structures: if

- if (condition){

-  do something

- }else{

-  do something else

- }

 
- if (condition){

- do something

- }else if (condition){

- do something different

- }else{

- do something different

- }

- else part is optional.

- There can be any number of else if condition but the else one has to be at the end.

## This is a valid if/else structure

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-1:10
if (x>3){
  y <-10
}else{
  y <-0
}

# So is this one
y <-if(x>3){
  10
}else{
  0
}

```

- The entire if else is about assigning a value to y.

## Control Structures - For loops

- This is the most common type of loop we are going to use in R.

- For loops take an interactor variable and assign it successive values from a sequence or vector.

- For loops are most commonly used for iterating over the elements of an object(list, vector etc).

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
for (i in 1:10){
  print(i)
}

```

- This loop takes the i variable and in each iteration of the loop gives it values 1,2,3,...,10 and then exits. i is the loop index.

- These three loops have the same behaviour.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-c("a","b","c","d")
for (i in 1:4){
  print(x[i])
}
for (i in seq_along(x)){
  print(x[i])
}
for (letter in x){
  print(letter)
}

```

- for loops are generated using curly brackets {}.

## Nested for loops

- for loops can be nested.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-matrix(1:6,2,3)
for (i in seq_len(nrow(x))){
  for (j in seq_len(ncol(x))){
    print(x[i,j])
  }
}

```

- Nesting beyond 2-3 levels is often very difficult to read/understand.

## Control Structures - While loops

- While loop is another major looping construct in R.

- while loop begin by;

1. Testing a condition.

2. If it is true, then they execute the loop body.

3. Once the loop body is executed, the condition is tested again, and so forth.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
count <-0
while (count < 10){
  print(count)
  count <-count + 1
}

```

- While loops can potentially result in;

1. infinite loops if not written properly.

2. Use with care.

- While takes a logical expression. While loops is useful because it makes things readable. Sometimes there will be more than one condition in the test.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
z <-5
while (z >=3&&z<=10){
  print(z)
  coin <-rbinom(1,1,0.5)
  if(coin==1){
    ## random walk
    z <-z+1
  }else{
    z <-z-1
  }
}

```

- Conditions are always evaluated from left to right.

- It is difficult to tell when the while construct will finish.

## Control Structures - Repeat,Next,Break

### Repeat

- Repeat initiates an infinite loop. These are not commonly used in statistical applications but they do have their uses. The only way to exit a repeat loop is to call break.


- x0 <-1

- tol <-exp(-8)

- repeat{

- x1 <-computeEstimate()

- if(abs(x1-x0) <-tol){

- break

- }else{

- x0 <-x1

- Uses;

1. Optimization

a. maximize a function.

b. find solution of some set of equations.

- computeEstimate is not a real function.

### Repeat

- The loop in the previous slide is a bit dangerous because there is no guarantee it will stop.

- Better to set a hard limit on the number of iterations (e.g. using a for loop) and then report whether convergence was achieved or not.

- It's difficult to predict this kind of loop so you can use a for loop to set a hard limit.

## Next, Return

- Next is used to skip an iteration of a loop.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
for (i in 1:100){
  if(i <=20){
    #skip the first 20 iterations
    next
  }
  #Do something else
}

```

- Return signals that a function should exit and return a given value.

- next skips a loop and break exits a loop.

## Control Structures

### Summary

- Control structures like if, while and for allow you to control the flow of an R program.

- Infinite loops should generally be avoided, even if they are theoretically correct.

- Control structures mentioned here are primarily useful for writing programs.

- For command line interactive work, the apply functions are more useful.

## Your First R function

- The first thing we are going to do is to put our function in a text file.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
add2 <-function(x,y){
  x + y
}
add2(3,5)

above10 <-function(x){
  use <-x>10
  x[use] #subset
}

above <-function(x,n){
  use <-x>n
  x[use] #subset
}
x <-1:20
above(x,12)

```

### function

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
column <-function(y){
  nc <-ncol(y)
  means <-numeric(nc)
  for (i in 1:nc){
    means[i] <-mean(y[,i])
  }
  means
}

```

- You can remove the NAs by using the code na.rm = TRUE.

## Functions - Part 1

- Functions in R are created using the function() directive and are stored as R objects just like anything else. In particular, they are R objects of "class" function.

- f <-function(argument){Do something interesting}

- Functions in R are first class objects which means that they can be treated much like any other R object.

- Importantly,

1. Functions can be passed as arguments to other functions.

2. Functions can be nested, so that you can define a function inside another function.

3. The return value of a function is the last expression in the function body to be evaluated.

## Function Arguments

- Functions have named arguments which potentially have default values.

- The formal arguments are the arguments included in the function definition.

- The formals function returns a list of all the formal arguments of a function.

- Not every function call in R makes use of all the formal arguments.

- Function arguments can be missing or might have default values.

- Formals function takes the function as inputs.

## Argument Matching

- R function arguments can be matched positionally or by name. So the following calls to sd are all equivalent.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
mydata <-rnorm(100)
sd(mydata)
sd(x=mydata)
sd(x=mydata,na.rm=FALSE)
sd(na.rm=FALSE,x=mydata)
sd(na.rm=FALSE,mydata)

```

- Even though it's legal, I don't recommend messing around with the order of the arguments too much, since it can lead to some confusion.

- When you name the arguments you don't have to put them in a special order.

- Name one argument and don't name the other argument will still give you your answer.

- You can mix positional matching with matching by name. When an argument is matched by name, it is "taken out" of the argument list and the remaining unnamed arguments are matched in the order that they are listed in the function definition.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
args(lm)

```

- Most of the time, named arguments are useful on the command line when you have a long argument list and you want to use the defaults for everything except for an argument near the end of the list.

- Named arguments also help if you can remember the name of the argument and not its position on the argument list (plotting is a good example).

- Function arguments can also be partially matched, which is useful for interactive work and not for programming on the command line.

- The order of operations when given an argument is;

1. Check for exact match for a named argument.

2. Check for a partial match.

3. Check for a positional match.

## Functions - Part2

### Defining a function

- f <-function(a,b=1,c=2,d=NULL).

- a has no default value. b, c and d have a default value.

- In addition to not specifying a default value, you can also set an argument value to NULL.

## Lazy Evaluation

- Arguments to functions are evaluated lazily, so they are evaluated only as needed. They don't use some of the variables (missing).

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
f <-function(a,b){
  a^2
}
f(2)

```

- This function never actually uses the argument b, so calling f(2) will not produce an error because the 2 gets positionally matched to a.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
f <-function(a,b){
  print(a)
  print(b)
}
f(4,5)

```

## The "..." Argument

- The ... argument indicate a variable number of arguments that are usually passed on to other functions. The ... is often used when extending another function and you don't want to copy the entire argument list of the original function.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
myplot <-function(x,y,type="l",...){
  plot(x,y,type=type,...)
}

```

- Generic functions use ... so that extra arguments can be passed to methods(more on this later).

- Generic functins dispatch methods for different types of data.

- The ... argument is also necessary when the number of arguments passed to the function cannot be known in advance.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
args(cat) #concatenate
args(paste)

```

## Arguments Coming After the "..." Argument

- One catch with ... is that any arguments that appear after ... on the argument list must be named explicitly and cannot be partially matched.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
args(paste)
paste("a","b",sep=":")
paste("a","b",se=":")
paste0("a","b")

```

- You can not use positional matching or partial matching for arguments that come after the dots.

## Scoping Rules - Symbol Binding

### A Diversion on Binding values to symbol

- How does R know which value to assign to which symbol? When I type;

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
lm <-function(x){x*x}
lm

```

- How does R know what value to assign to the symbol lm? Why doesn't it give it the value of lm that is in the stats package?

- When R tries to bind a value to a symbol, it searches through a series of environments to find the appropriate value.

- When you are working on the command line and need to retrieve the value of an R object, the order is roughly;

1. Search the global environment for a symbol name matching the one requested.

2. Search the names spaces of each of the packages on the search list.

3. The search list can be found by using the search function.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
search()

```

## Binding Values to Symbol

- The global environment or the user's workspace is always the first element of the search list and the base package is always the last.

- The order of the packages on the search list matters!

- User's can configure which packages get loaded on startup so you cannot assume that there will be a set list of packages available.

- When a user loads a package with library, the names space of that package gets put in position 2 of the search list (by default) and everything else gets shifted down the list.

- Note that R has separate namespaces for functions and non-functions so it's possible to have an object named C and a function named C.

## Scoping Rules for R

- The scoping rules for R are the main features that make it different from the original S language.

- The scoping rules determine how a value is associated with a free variable in a function.

- R uses lexical scoping or static scoping. A common alternative is dynamic scoping.

- Related to the scoping rules is how R uses the search list to bind a value to a symbol.

- Lexical scoping turns out to be particularly useful for simplifying statistical computations.

## Lexical Scoping

- Consider the following function.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
f <-function(x,y){
  x^2+y/z
}

```

- This function has two(2) formal arguments x and y. In the body of the function there is another symbol z. In this case z is called a free variable. The scoping rules of a language determine how values are assigned to free variables. Free variables are not formal arguments and are not local variables(assigned inside the function body).

### Lexical Scoping

- Lexical scoping in R means that the values of free variables are searched for in the environment in which the function was defined.

### What is an Environment?

- An environment is a collection of (symbol,value) pairs i.e. x is a symbol and 3.14 might be its value.

- Every environment has a parent environment. It is possible for an environment to have multiple "children".

- The only environment without a parent is the empty environment.

- A function + an environment = a closure or function closure.

### Lexical Scoping

- Searching for the value for a free variable.

- If the value of a symbol is not found in the environment in which a function was defined, then the search is continued in the parent environment.

- The search continues down the sequence of parent environments until we hit the top-level environment. This is usually the global environment(workspace) or the namespace of a package.

- After the top level environment, the search continues down the search list until we hit the empty environment.

- If a value for a given symbol cannot be found once the empty environment is arrived at, then an error is thrown.

- It is possible to define a function outside the global environment.

- After the base package we hit the empty environment.

## R Scoping Rules

### Lexical Scoping

- Why does all this matter?

- Typically a function is defined in the global environment, so that the values of free variables are just found in the user's workspace.

- This behaviour is logical for most people and is usually the right thing to do.

- However in R, you can have functions defined inside other functions. Languages like C don't let you do this.

- Now things get interesting - In this case the environment in which a function is defined is the body of another function!

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
make.power <-function(n){
  pow <-function(x){
    x^n
  }
  pow
}
# This function returns another function as its value
cube <-make.power(3)
square <-make.power(2)
cube(3)
square(3)

```

## Exploring a function Closure

### What's in a function's environment?

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
ls(environment(cube))
get("n",environment(cube))
ls(environment(square))
get("n",environment(square))

```

## Lexical vs. Dynamic Scoping

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
y <-10
f <-function(x){
  y <-2
  y^2+g(x)
}
g <-function(x){
  x*y
}
f(3) #what is the value of f(3)

```

- In the f function y is a free variable and g is also a free variable.

- In the g function y is a free variable.

- A free variable is a variable that is not defined in a function.

### Lexical vs Dynamic Scoping

- With lexical scoping the value of y in the function g is looked up in the environment in which the function was defined, in this case the global environment, so the value of y is 10.

- With dynamic scoping, the value of y is looked up in the environment from which the function was called (sometimes referred to as the calling environment).

- In R the calling environment is known as the parent frame. So the value of y will be 2.

- When a function is defined in the global environment and is subsequently called from the global environment, then the defining environment and the calling environment are the same. This can sometimes give the appearance of dynamic scoping when it even does not exist.

- In this case dynamic scoping and lexical scoping are the same.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
g <-function(x){
  a <-3
  y <-3
  x+a+y
}
g(2)

```

## Other Languages

- There are a number of languages that support lexical scoping.

1. Scheme.

2. Perl.

3. Python.

4. common Lisp(all languages converge to Lisp).

- The acronym is CLiPPS i.e. Common Lisp Perl Python Scheme.

## Consequences of Lexical Scoping in R

- In R, all objects must be stored in memory.

- All functions must carry a pointer to their respective defining environments, which could be anywhere.

- In S-PLUS, free variables are always looked up in the global workspace, so everything can be stored on the disk because the "defining environment" of all functions is the same.

## Scoping Rules - Optimization Example (Optional)

### Application: Optimization

- Why is any of this information used?

- Optimization routines in R: Like optim, nlm, and optimize require you to pass a function whose argument is a vector of parameters (e.g. log-likelihood).

- However, an object function might depend on a host of other things besides its parameters (like data).

- When writing software which does optimization, it may be desirable to allow the user to hold certain parameters fixed.

### Maximizing a Normal Likelihood

- Write a "Constructor" function.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
make.NegLogLik <-function(data,fixed=c(FALSE,FALSE)){
  params <-fixed
  function(p){
    params[!fixed] <-p
    mu <-params[1]
    sigma <-params[2]
    a <- -0.5*Length(data)*log(2*pi*sigma^2)
    b <- -0.5*sum((data-mu)^2)/(sigma^2)
    -(a+b)
  }
}

```

- Optimization functions in R minimize functions, so you need to use the negative log-likelihood.

## Maximize a Normal Likelihood

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(1)
normals <-rnorm(100,1,2)
nLL <-make.NegLogLik(normals)
nLL
function(p){
  params[!fixed] <-p
  mu <-params[1]
  sigma <-params[2]
  a <- -0.5*length(data)*log(2*pi*sigma^2)
  b <- -0.5*sum((data-mu)^2)/(sigma^2)
  -(a+b)
}

ls(environment(nLL))

```

## Estimating Parameters

- optim(c(mu=0,sigma=1),nLL)$par

- nLL <-make.NegLogLik(normals,c(FALSE,2))

- optimize(nLL,c(-1,3))$minimum

- Fixing u=1

- nLL <-make.NegLogLik(normals,c(1,FALSE))

- optimize(nLL,c(1e^-6,10))$minimum

## Plotting the Likelihood

- nLL <-make.NegLogLik(normals,c(1,FALSE))

- x <-seq(1.7,1.9,len=100)

- y <-sapply(x,nLL)

- plot(x,exp(-(y-min(y))),type="l")

- nLL <-make.NegLogLik(normals,c(FALSE,2))

- x <-seq(0.5,1.5,len=100)

- y <-sapply(x,nLL)

- plot(x,exp(-(y-min(y))),type="l")

## Lexical Scoping Summary

- Objective functions can be "built" which contain all of the necessary data for evaluating the function.

- No need to carry around long argument lists - using for interactive and exploratory work.

- Code can be simplified and cleaned up.

- Reference: Robert Gentleman and Ross Ihaka(2000). "Lexical Scope and Statistical Computing" JCGS,9,491-508.

## Coding Standards(very important)

1. Always use text files/text editor(save as text files). Reason: So that people can use any editor to access and read your code.

2. Indent your code - it makes it readable.

3. Limit the width of your code(80 columns?). This standard moves hand in hand with indenting.

4. Limit the length of individual functions.

## Change Indenting Scheme

- Rstudio icon.

- Preferences menu.

- Code editing.

- Change tab to 4.

- Margin column 80.

## Identing(recommendations)

- Indenting improves readability.

- Fixing line length (80 columns) prevents lots of nesting and very long functions.

1. lots of nesting.

2. very long functions.

- Suggested: Indents of 4 spaces at minimum; 8 spaces ideal. 8 spaces prevents you from making basic mistakes with code readability. It helps you write your code efficiently.

## Coding Standards for R

### Advantages

- First of all it's nice to have your function written on a single page. You don't have to scroll back and forth(endlessly).

- It helps you trace your bugs(debugger, traceback, profiler). It helps for debugging and profiling.









