---
title: "ggplot2 Graphics"
author: "David Asare Kumi"
date: "June 10, 2019"
output: html_document
---


##What is ggplot2?

- An implementation of the grammar of graphics by Leland Wilkinson.

- Written by Hadley Wickham (while he was a graduate student at Iowa State University).

- A "third" graphics system for R (along with base and lattice).

- Available from CRAN via install.packages().

- Website: http://ggplot2.org (better documentation).

- Grammar of graphics represents an abstraction of graphics ideas/objects.

- Think "verb","noun","adjective" for graphics.

- Allows for a theory of graphics on which to build new graphics and graphics objects.

- According to Hadley Wickham, he wants to "shorten the distance from mind to page".


##The Basics: qplot()

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
dim(mpg)
str(mpg)
#some summary tables
summary(mpg)
table(mpg$cyl)
table(mpg$class)
table(mpg$drv)
table(mpg$fl)
#scatterplot
qplot(displ,hwy,data=mpg)
#specify color=drv(a factor of categorical variable)
qplot(displ,hwy,data=mpg,color=drv,main="Scatterplot of Displ and Hwy")
#adding a geom
qplot(displ,hwy,data=mpg,geom=c("point","smooth"))
#histogram
qplot(hwy,data=mpg,geom="histogram")
qplot(hwy,data=mpg,fill=drv)
qplot(hwy,data=mpg,facets=.~drv)
#density
qplot(hwy,data=mpg,geom="density")
qplot(hwy,data=mpg,geom="density",color=drv)
qplot(hwy,data=mpg,geom="density",facets=.~drv)

```

##Facets

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
qplot(displ,hwy,data=mpg,facets=.~drv)+geom_smooth(method="lm")
qplot(displ,hwy,data=mpg,facets=.~drv)+geom_smooth(method="lm",se=FALSE)

```

##Scatterplots

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
qplot(displ,hwy,data=mpg)
qplot(displ,hwy,data=mpg,shape=drv)
qplot(displ,hwy,data=mpg,color=drv)+geom_smooth(method="lm",se=FALSE)
#facets are used in splitting up the data and also for conditioning on factor variables
qplot(displ,hwy,data=mpg,facets=.~drv)+geom_smooth(method="lm",se=FALSE)

```

##Building Plots with ggplot2

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
g <-ggplot(mpg,aes(displ,hwy))
g1<-geom_point(size=2,color="red")
g2<-facet_grid(.~drv)
g3<-geom_smooth(method="lm",se=FALSE)
g4<-labs(title="Scatterplot of Displ and Hwy conditioned on Drv")

g #This gives you the plot area
g+g1 #This gives the plot area plus the plotted points
g+g1+g2 #This splits the entire graph according to the factor variable drv
g+g1+g2+g3 #This fits the straight line on the splitted graphs
g+g1+g2+g3+g4 #This adds the title to the final plot.

```
