---
title: "Lattice Plotting System(Part1)"
author: "David Asare Kumi"
date: "June 10, 2019"
output: html_document
---


##Lattice Plots

- Different from base plotting system.

- Useful for plotting high dimensional data.

- Using for making many many plots at (panel plots) once.

##Functions of the Lattice Plotting System

The lattice plotting system is implemented using the following packages.

- Lattice: Contains code for producing Trellis graphics which are independent of the base graphics system; includes functions like xyplots,bwplots,levelplots.

- Grid: implements a different graphing system independent of the base system; the lattice package builds on top of grid. We seldom call functions from the grid package directly.

- The lattice plotting system does not have a two phase aspect with separate plotting and annotation like in base plotting.

- All plotting/annotation is done at once with a single function call.

- Some of the main lattice functions include;

1. xyplot; this is the main function for creating scatterplots.

2. bwplot; box and whiskers plots ("boxplots").

3. histogram; histograms.

4. stripplot; like a boxplot but with actual points.

5. dotplot; plot dots on "violin strings".

6. Splom; scatterplot matrix - like pairs in base plotting system.

7. levelplot, contourplot; for plotting "image" data.


##Simple lattice plot

```{r,echo=TRUE,message=FALSE,warning=FALSE,fig.width=10}
library(datasets)
library(lattice)
data(airquality)
names(airquality)
xyplot(Ozone~Wind,data=airquality)

#lattice plot involving factor variable
library(lattice)
library(datasets)
data(airquality)
str(airquality)
Month <-as.factor(airquality$Month)
class(Month)
xyplot(Ozone~Wind|Month,data=airquality,layout=c(5,1))

```

##Lattice Plotting System (Part 2)

- Lattice functions have a panel function which controls what happens inside each panel of the plot.

- The lattice package comes with default panel functions, but you can supply your own if you want to customize what happens in each panel.

- Panel functions receive the x/y coordinates of the data points in their panel (along with any optional arguments).


##Custom Panel function

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(lattice)
library(datasets)
data(airquality)
names(airquality)
Month <-as.factor(airquality$Month)
x <-airquality$Wind
y <-airquality$Ozone
xyplot(y~x|Month,panel=function(x,y,...){
  panel.xyplot(x,y,...)
  panel.lmline(x,y,col="red")
})

```

