---
title: "The Structure Function"
author: "David Asare Kumi"
date: "11/1/2019"
output: html_document
---


## Str Function

- str compactly display the internal structure of an R object.

- A diagnostic function and an alternative to "summary".

- It is especially well suited to compactly display the (abbreviated) contents of (possibly nested) lists.

- Roughly one line per basic object.

## What's in this object

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
str(str)
str(lm)
x <-rnorm(100,2,4)
summary(x)
str(x)
f <-gl(40,10)
str(f)
summary(f)

```

- gl function: Generate factors by specifying the pattern of their levels. Simply generate factor levels.

- Usage: gl(n, k, length = n*k, labels = seq_len(n), ordered = FALSE)

- Arguments

- n: an integer giving the number of levels.

- k: an integer giving the number of replications.

- length: an integer giving the length of the result.

- labels: an optional vector of labels for the resulting factor levels.

- ordered: a logical indicating whether the result should be ordered or not.

- Value: The result has levels from 1 to n with each value replicated in groups of length k out to a total length of length. gl is modelled on the GLIM function of the same name.

### Example

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
# First control, then treatment:
gl(2, 8, labels = c("Control", "Treat"))
# 20 alternating 1s and 2s
gl(2, 1, 20)
# alternating pairs of 1s and 2s
gl(2, 2, 20)

```

### Example

-str is very useful for quick examination of your dataset. Str is the most useful function in R.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
data(airquality)
str(airquality)
s <-split(airquality,airquality$Month)
s
str(s)

```

## Simulation - Generating Random Numbers

- Functions for probability distributions in R.

- rnorm: generate random number variates with a given mean and standard deviation.

- dnorm: evaluate the normal probability density (with a given mean and sd) at a point (or vector of points).

- pnorm: evaluate the cumulative distribution function for a Normal distribution.

- rpois: generate random Poisson variates with a given rate.

- Probability distribution functions usually have four functions associated with them. The functions are prefixed with a;

1. d for density.

2. r for random number generation.

3. p for cumulative distribution.

4. q for quantile function.

- Every distribution has these four functions (d,r,p,q).

- Working with the Normal distributions requires using these four functions.

- dnorm(x,mean=0,sd=1,log=FALSE).

- pnorm(q,mean=0,sd=1,lower.tail=TRUE,log.p=FALSE).

- qnorm(p,mean=0,sd=1,lower.tail=TRUE,log.p=FALSE).

- rnorm(n,mean=o,sd=1).

- If phi is the cumulative distribution function for a standard normal distribution, then pnorm(q)=phi(q) and qnorm(p)=phi inverse (p).

## Generating Random Numbers

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-rnorm(10)
x
x1 <-rnorm(10,20,2)
x1
summary(x1)

```

- Setting the random number seed with set.seed ensures reproducibility.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(1)
rnorm(5)
rnorm(5)

```

- Always set the random number seed when conducting a simulation!

- The numbers are not actually random. They are called pseudo random numbers.

- When you do a simulation you always set the seed so that you can go back and generate the same results.

## Generating Poisson Data

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
rpois(10,1) # rate=mean=1
rpois(10,2)
rpois(10,20)
ppois(2,2) # cumulative distribution
ppois(4,2) # p(x <= 4)
ppois(6,2) # p(x <= 6)

```

- The cumulative distribution allows you to evaluate these probabilities.

## Simulation - Simulating a linear model

### Generating Random Numbers from a linear model.

- Suppose we want to simulate from the following linear model. Y = Bo + B1X + E where E~N(0,2^2).

- Assume x~N(0,1), Bo=0.5 and B1=2.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(20)
x <-rnorm(100)
e <-rnorm(100,0,2)
y <-0.5 + 2*x + e
summary(y)
plot(x,y,cex=1,pch=20,col="red")

```

- What if x is binary?

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(10)
x <-rbinom(100,1,0.5)
e <-rnorm(100,0,2)
y <-0.5 + 2*x + e
summary(y)
plot(x,y,cex=2,col="blue")

```

- x binary: may be x represents gender. Here we use the binomial distribution. rbinom(100,1,0.5) means n=1 and p=0.5.

## Generating Random Numbers from a generalized linear model

- Suppose we want to simulate from a poisson model where Y~Poisson(U), logU = Bo + B1x and Bo=0.5 and B1=0.3.

- We need to use the rpois function for this.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(1)
x <-rnorm(100)
log.U <-0.5 + 0.3*x
y <-rpois(100,exp(log.U))
summary(y)
plot(x,y,cex=1,pch=20,col="red")

```

- In this instance, the error is not going to be normal. It is going to be poisson. There is a linear relationship between x and y. This looks like poisson data.

## Simulation - Random Sampling

- The sample function draws randomly from a specified set of (scalar) objects allowing you to sample from arbitrary distribution.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
set.seed(1)
sample(1:10,4)
sample(letters,5)
sample(1:10) # permutation
sample(1:10,replace=TRUE) # sample with replacement

```

## Simulation

### Summary

- Drawing samples from specific probability distributions can be done with r * function.

- Standard distributions are built in: Normal, Poisson, Binomial, Exponential, Gamma etc.

- The sample function can be used to draw random samples from arbitrary vectors.

- Setting the random number generator seed via set.seed is critical for reproducibility.
