---
title: "Loop Functions - lapply"
author: "David Asare Kumi"
date: "10/29/2019"
output: html_document
---


## Looping on the command line

- Writing for, while loops is useful when programming but not particularly easy when working interactively on the command line. There are some functions which implement looping to make life easier.

- Lapply: loop over a list and evaluate a function on each element.

- Sapply: same as lapply but try to simplify the results.

- Apply: A function over the margins of an array.

- Tapply: Apply a function over subsets of a vector.

- Mapply: Multivariate version of lapply.

- An auxiliary function "split" is also useful, particularly in conjunction with lapply.

## Lapply

- lapply takes 3 arguments;

1. a list x.

2. a function (or the name of a function)FUN.

3. Other arguments via its ... argument.

- If x is not a list, it will be coerced to a list using as.list.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
function(x,FUN,...){
  FUN <-match.fun(FUN)
  if(!is.vector(x)||is.object(x))
    x <-as.list(x)
  .Internal(lapply(x,FUN))
}

```

- The actual looping is done internally in c code.

- If it is not possible to coerce x into a list you will get an error.

- lapply always returns a list, regardless of the class of the input. A list can contain any kind of object like matrices, vectors, data frame etc.

## Example

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-list(a=1:5,b=rnorm(10))
lapply(x,mean)

```

- The list is the same name as the original list a and b.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-list(a=1:4,b=rnorm(10),c=rnorm(20,1),d=rnorm(100,5))
lapply(x,mean)
x <-1:4
lapply(x,runif)
x <-1:4
lapply(x,runif,min=0,max=10)

```

- runif: This is uniform random variables.

- 0, 1 is the default for uniform random variable.

- The min = 0 and max = 10 are run within the ... argument of lapply.

- lapply and friends (associated functions) make heavy use of anonymous functions.

- Anonymous functions are functions that do not have a name so you don't assign them a name but you can generate them on the fly.

- Suppose you want to extract the first column of the matrix below.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-list(a=matrix(1:4,2,2),b=matrix(1:6,3,2))
x
#anonymous function for extracting the first column of each matrix
lapply(x,function(elt)elt[,1])
sapply(x,function(elt)elt[,1])

```

- This function does not exist except within the context of lapply and after the lapply function it is finished. This is an anonymous function.

- sapply also returns a list when we use the anonymous function.

## Sapply

- This is the variant of lapply(alternate).

- Sapply will try to simplify the result of lapply if possible.

- If the result is a list where every element is length 1, then a vector is returned.

- If the result is a list where every element is a vector of the same length(>1), a matrix is returned.

- If it can't figure things out, a list is returned.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-list(a=1:4,b=rnorm(10),c=rnorm(20,1),d=rnorm(100,5))
lapply(x,mean)
sapply(x,mean)

```

- Note that mean(x) will return an error because mean is not meant to be applied to list.

## Apply

- Apply function is used to evaluate a function(often an anonymous one) over the margins of an array.

- It is most often used to apply a function to the rows or columns of a matrix.

- It can be used with general arrays, e.g. taking the average of an array of matrices.

- It is not really faster than writing a loop, but it works in one line!

- We always use apply because it involves less typing. Less typing is always better because good programs are always lazy.

- Apply is very useful on the command line because you want to do little typing.

## How does apply work?

- str(apply)= args(apply). function(x,MARGIN,FUN,...).

- x is an array.

- MARGIN is an integer vector indicating which margins should be "retained".

- FUN is a function to be applied.

- ... is for other arguments to be passed to FUN.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-matrix(rnorm(200),20,10)
apply(x,2,mean) #This gives the mean of each of the 10 columns
apply(x,1,mean) #This gives the mean of each of the 20 rows
apply(x,2,sum) # This gives the sum of each of the 10 columns
apply(x,1,sum) # This gives the sum of each of the 20 rows.

```

- It is important to note that, 2 stands for columns and 1 stands for rows.

## Col/Row Sums and Means

- For sums and means of matrix dimensions, we have some shortcuts.

- rowSums = apply(x,1,sum).

- rowMeans = apply(x,1,mean).

- colSums = apply(x,2,sum).

- colMeans = apply(x,2,mean).

- The shortcut functions are much faster, but you wouldn't notice unless you are using a large matrix.

## Other Ways to Apply

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-matrix(rnorm(200),20,10)
apply(x,1,quantile,probs=c(0.25,0.75))

```

- Average matrix in an array.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
a <-array(rnorm(2*2*10),c(2,2,10)) # 2rows*1columns*10dims
apply(a,c(1,2),mean)
rowMeans(a,dims=2)

```

## Loop Functions - mapply

- mapply is a multivariate apply of sorts which applies a function in parallel over a set of arguments.

- str(mapply). function(FUN,...,MoreArgs=NULL,SIMPLIFY=TRUE,USE.NAMES=TRUE).

- FUN is a function to apply.

- ... contains arguments to apply over.

- MoreArgs is a list of other arguments to FUN.

- SIMPLIFY indicates whether the result should be simplified.

- lapply, sapply, tapply, apply functions over a single list. With mapply it can take multiple list and applies the function in parallel.

## Example

- The following is tedious to type.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
list(rep(1,4),rep(2,3),rep(3,2),rep(4,1))
# Instead we can do 
mapply(rep,1:4,4:1)

```

- mapply can be used to apply functions to multiple sets of arguments.

## Vectorizing a function

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
noise <-function(n,mean,sd){
  rnorm(n,mean,sd)
}
noise(5,1,2)
noise(1:5,1:5,2)
# This function does not really work properly

```

- When we use the mapply function we get a more efficient results.

## Instant Vectorization

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
mapply(noise,1:5,1:5,2)

```

- Which is the same as:

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
list(noise(1,1,2),noise(2,2,2),noise(3,3,2),noise(4,4,2),noise(5,5,2))

```

- Here we fix the sd to be 2 and be changing n and the mean.

- This is how we instantly vectorize a function that doesn't allow for vector arguments/vector inputs.

## Loop Functions - Tapply

- tapply is used to apply a function over subsets of a vector. I don't know why it is called tapply.

- str(tapply). function(x,INDEX,FUN=NULL,...,simplify=TRUE).

- x is a vector.

- INDEX is a factor or a list of factors(or else they are coerced to factors).

- FUN is a function to be applied.

- ... contains other arguments to be passed to FUN.

- simplify, should we simplify the results?

- Index could be male or female (factor variable).

## tapply

- Take group means.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-c(rnorm(10),runif(10),rnorm(10,1))
f <-gl(3,10)
f
tapply(x,f,mean)
#Take group means without simplification, you get a list
tapply(x,f,mean,simplify=FALSE)
#Find group ranges
tapply(x,f,range) #List with a vector of length 2

```

## Loop Functions - Split

- Split takes a vector or other objects and splits it into groups determined by a factor or list of factors.

- str(split). function(x,f,drop=FALSE,...).

- x is a vector (or list) or data frame.

- f is a factor (or coerced to one) or a list of factors.

- drop indicates whether empty factor levels should be dropped.

- Split is not a loop function but it is a very handy function that can be used along side lapply, tapply etc.

- After splitting, you can then use the other loop functions.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-c(rnorm(10),runif(10),rnorm(10,1))
f <-gl(3,10)
split(x,f) #split according to the factor variable.

```

- Split always returns the list back.

- A common idiom is split followed by an lapply.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-c(rnorm(10),runif(10),rnorm(10,1))
f <-gl(3,10)
Sp <-split(x,f)
lapply(Sp,mean)
#alternatively
lapply(split(x,f),mean)
tapply(x,f,mean) #use tapply function which is more compact to derive the same results.

```

## Splitting a Data Frame

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
data(airquality)
head(airquality)
tail(airquality)
dim(airquality)
summary(airquality)
s <-split(airquality,airquality$Month)
lapply(s,function(x)colMeans(x[,c("Ozone","Solar.R","Wind")],na.rm=TRUE))
#The factor we use here is the month variable

```

- The anonymous function here is the colMeans.

- Each element of the list is a vector of length 3.

- We can also call sapply here.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
sapply(s,function(x)colMeans(x[,c("Ozone","Solar.R","Wind")],na.rm=TRUE))

```

- The NA's are removed. When you have the NA's in the data, you cannot calculate the means.

- Split is a very handy function for splitting arbitrary objects according to the levels and factor and then apply any of the loop functions (lapply/sapply).

## Splitting on more than one level:

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
x <-rnorm(10)
f1 <-gl(2,5)
f2 <-gl(5,2)
f1
f2
interaction(f1,f2) #interactions can create empty levels
str(split(x,list(f1,f2)))
str(split(x,list(f1,f2),drop=TRUE))
split(x,list(f1,f2),drop=TRUE)

```

- When you use the split function, you don't have to use the interaction function, you just have to pass in the list with the factors. This also returns a list.

- To drop the empty levels, the split function has an argument drop.



