---
title: "Coding Standards in R"
author: "David Asare Kumi"
date: "9/2/2019"
output: html_document
---


## Coding Standards in R

- Useful

1. Help you make your code readable.

2. Allow you and other people understand what's going on.

- Principles; always write your code in R using;

1. Always use text files/test editor; This can be read by any basic editor. Every one can access the code when its a text.

- Indent your code; This is the idea that different blocks of codes should be spaced to the right a little bit more than others so that you can see how the codes flow.

- Limiting the width of your codes (80 columns?) i.e. a recommendation.

- The editor in Rstudio is a text editor. To change indenting scheme;

1. go to preferences menu

2. go to code editing - change to 4(tab width).

3. margin column is set at 80 select all with command A and then command i. The structure of the program is much more obvious.

4. For indenting choose between 4 to 8 spaces. This prevents you from making basic mistakes. Indenting also makes your code readable.

- Limit the length of your (individual) functions.

1. limit it to one basic activity (function).

## Advantage

- Write your function on a page.

- Here you can use traceback, profiler more efficiently.

- Helps in debugging.

- Limiting the size of your function makes it readable.

## Markdown

- Useful for;

1. Creating documents.

2. Easy to integrate with R code and other languages.

## What is Markdown?

- Markdown is a text-to-HTML conversion tool for web writers. Markdown allows you to write using an easy-to-read, easy-to-write plain text format, then convert it to structurally valid XHTML (or HTML).

- John Gruber is the creator of Markdown.

## Markdown Syntax

- Italics; *This text will appear italicized!*

- Bold; **This text will appear bold!**

## Headings

- Secondary heading; ## This is a secondary heading.

- Tertiary heading; ### This is a tertiary heading.

- Primary heading; # The single harsh is the primary heading.

## Create Simple Unordered Lists

- first item in list.

- second item in list.

- third item in list.

## Ordered Lists

1. first item in list.

2. second item in list.

3. third item in list.

## Links

- There are two ways to create links.

1. [Johns Hopkins Blomberg School of Public Health](http://www.jhsph.edu/)

[DownloadR](http://www.r-project.org/)

[RStudio](http://www.rstudio.com/)

2. Another way to create a link. Advanced linking. I spend so much time reading [Rbloggers][1] and [Simply Statistics][2]!

[1]: http://www.r-bloggers.com/"Rbloggers"

[2]: http://simplystatistics.org/"Simply Statistics"

- This type of linking makes your text a little bit cleaner to read and easier to read.

## New Lines

- New lines require a double space after the end of a line.

- First line  Second line.

## Markdown Resources

- The Official Markdown Documentation.

- Github's Markdown Guide.

## R Markdown

- The idea is that we want to have R language integrated with a markdown language.

## What is Markdown?

- Created by John Gruber and Aaron Swartz.

- A simplified version of "markup" languages.

- Allows one to focus on writing as opposed to formatting.

- Simple/minimal intuitive formatting elements.

- Easily converted to valid HTML (and other formats) using existing tools.

- Complete information is available at http://daringfireball.net/projects/markdown/

- Some background information at http://daringfireball.net/2004/03/dive_into_markdown


## What is RMarkdown?

- R markdown is the integration of R code with markdown.

- Allows one to create document containing "live" R code.

- R code is evaluated as part of the processing of the markdown.

- Results from R code are inserted into markdown document.

- A core tool in literate statistical programming.

- R markdown has .rmd extension.

- R markdown can be converted to standard markdown using the knitr package in R.

- Markdown can be converted to HTML using markdown package in R.

- Any basic text editor can be used to create a markdown document; no special editing tools needed.

- The Rmarkdown ---> markdown ---> HTML workflow can be easily managed using RStudio (but not required).

- These slides were written in R markdown and converted to slides using the slidify package.

- You do not edit the markdown file and the HTML file. The only file you can edit is the Rmarkdown file.

## Rmarkdown demonstration

- RStudio

- Menu

- Rmarkdown

- Save as Rmarkdown demo.Rmd

- knitr; this converts markdown to html

## Load some data

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
data(airquality)
summary(airquality)

```

- Let's first make a pair.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
pairs(airquality)

```

- Here's a regression model of Ozone on Wind, Solar.R, and Temperature.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(stats)
fit <-lm(Ozone~Wind+Solar.R+Temp,data=airquality)
summary(fit) # always save and knitr

```

## knitr (part 1)

- Literate statistical programming with knitr.

### Problems, Problems

- Authors must undertake considerable effort to put data/results on the web.

- Readers must download data/results individually and piece together which data go with which code sections etc.

- Authors/readers must manually interact with websites.

- There is no single document to integrate data analysis with textual representations i.e. data, code, and text are not linked.

## Literate Statistical Programming

- Original idea comes from Don Knuth.

- An article is a stream of text and code.

- Analysis code is divided into text and code "chunks".

- Presentation code formats results (tables,figures etc).

- Article text explains what is going on.

- Literate programs are weaved to produce human readable documents and tangled to produce machine readable documents.

- Human readable documents are text files and machine readable documents are code files.

- Literate programming is a general concept. We need;

1. A documentation language.

2. A programming language.

- The original sweave system developed by Friedrich Leisch used LaTeX and R.

- knitr supports a variety of documentation languages.

- LaTeX is a documentation language and R is a programming language.

### The basic question is how do I make my work reproducible

- Decide to do it (ideally from the start).

- Keep track of things, perhaps with a version control system to track snapshot/changes.

- Use software whose operations can be coded.

- Don't save output.

- Save data in non-proprietary formats.

1. formats whose layout is not publicly known.

## Literate Programming

### Pros

- Text and code all in one place, logical order.

- Data results automatically updated to reflect external changes.

- Code is live -- automatic "regression test" when building a document.

### Cons

- Text and code all in one place; can make documents difficult to read, especially if there is a lot of code.

- Can substantially slow down processing of documents (although there are tools to help).

## Knitr (part 2)

### What is knitr?

- An R package written by Yihui Xie (while he was a grad student at Iowa state). The package is available on CRAN.

- Supports Rmarkdown, LaTeX, and HTML as documentation languages.

- Can export to HTML, pdf.

- Built right into RStudio for your convenience.

## Requirements

- A recent version of R.

- A text editor (the one that comes with RStudio) is okay.

- Some support packages also available on CRAN.

- Some knowledge of Markdown, LaTeX, or HTML.

- We will use markdown here.

## What is Markdown?

- A simplified version of "markup" languages.

- No special editor required.

- Simple, intuitive formatting elements.

- Complete information available at; http://google/MUt9i5

### What is knitr good for?

- Manuals.

- Short/Medium-length technical documents.

- Tutorials.

- Reports (especially if generated periodically).

- Data preprocessing documents/summaries.



### What is knitr not good for?

- Very long research articles.

- Complex time - consuming computations.

- Documents that require precise formatting.

- Formatting text in knitr is very adhoc.

## knitr (part 3)

- My first knitr document.

1. Click on file.

2. Click on new file.

3. Click on Rmarkdown and then;

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
# means you want to start a code chunk
set.seed(1)
x <-rnorm(100)
mean(x)
# close the chunk

```

4. Push the knit HTML button.

5. Note; you can name your code chunks.

## If you are not in RStudio

### More complicated way

- library(knitr)

- setwd(<working directory>).

- knit2html("document.Rmd").

- browseURL("document.html").

### A few Notes

- knitr will fill a new document with filler text;delete it.

- Code chunks begin with ```{r} and end with ```.

- All R code goes in between these markers.

- Code chunks can have names, which is useful when we start making graphics.

- By default, code in a code chunk is echoed, as will the results of the computation (if there are results to print).

## knitr (part 4)

- Processing of knitr documents (what happens under the hood).

1. You write the Rmarkdown document(.Rmd).

2. knitr produces a Markdown document(.md).

3. knitr converts the markdown document into HTML (by default).

4. .Rmd-->.md-->.html

5. You should not edit or save the .md or .html documents until you are finished. Only edit the Rmarkdown file.

## Note

1. when you specify echo=FALSE in the r chunk, the r code that produces the output will not show up.

2. You can also hide the results by specifying results="hide".

## Inline Text computations

- time <- format(Sys.time(),"%a%b%d %X%Y") gives the current date and time.

## Incorporating Graphics

- Let's first simulate some data and plot the data.

```{r,simulatesomedata,echo=TRUE,fig.height=4}
x <-rnorm(100)
y <-x + rnorm(100,sd=0.5)
par(mar=c(5,4,2,1),las=1)
plot(x,y,main="My Simulated Data",cex=1,pch=20,col="red")

```

## Making Tables with Xtable

```{r fitmodel}
library(datasets)
data(airquality)
fit <-lm(Ozone~Wind+Temp+Solar.R,data=airquality)

```

- Here is a table of regression coefficients.

```{r showtable,results="asis"}
library(xtable)
xt <-xtable(summary(fit))
print(xt,type="html")

```

## Setting Global Options

- Sometimes we want to set options for every code chunk that are different from the defaults.

- For example, we may want to suppress all code echoing and results output.

- We have to write some code to set these global options.

```{r setoptions,echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,results="hide")

```

### Some common Options

- Output

1. results:"asis","hide"

2. echo: TRUE, FALSE

- Figures

1. fig.height: numeric

2. fig.width: numeric

- asis: Show the raw results, don't process (post) no html processing.

## Caching Computations

- This is a strategy used to avoid waiting for a long time.

- What if one chunk takes a long time to run?

- All chunks have to be re-computed every time you re-knit the file.

- The cache=TRUE option can be set on a chunk-by-chunk basis to store results of computation.

- After the first run, results are loaded from cache.

## Caching Caveats

- If the data or code (or anything external) changes, you need to re-run the cached code chunks.

- Dependencies are not checked explicitly.

- Chunks with significant side effects may not be cacheable.

## Summary

- Literate statistical programming can be a useful way to put text, code, data, output all in one document.

- knitr is a powerful tool for integrating code and text in a simple document format.

- To create an Rmarkdown  report, open a plain text file and save it with the extention .Rmd.

- You can open a plain text file in your scripts editor by clicking; File>new file>textfile in Rstudio toolbar. Be sure to save the file with the extension .Rmd.

- In CRAN, install.packages("rmarkdown") then library(rmarkdown).

- We knit and convert documents at the same time.

- You can manually render an Rmarkdown file with rmarkdown::render().

- Render: knit and convert as an html file.

- echo=FALSE, is added to the code chunk to prevent printing of the R code that generated the plot.

- In practice, you don't need to call rmarkdown::render(). You can use a button in the RStudio IDE to render your report.

## Reproducible Research - Assessment 1 (Project)

- library(data.table)

- library(ggplot2)

- fileUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

- download.file(fileUrl,destfile=paste0(getwd(),"repdata%2Fdata%2FFactivity.zip"),method="Curl")

- unzip("repdata%2Fdata%2Factivity.zip",exdir="data")

### Reading CSV data into data.table

- activityDT <-data.table::fread(input="data/activity.csv")

### What is mean total number of steps taken per day?

- Calculate the total number of steps taken per day.

- TotalSteps <-activityDT[,c(lapply(.SD,sum,na.rm=FALSE)),.SDcols=c("steps"),by=.(date)]

- head(TotalSteps)

### Make a histogram of the total number of steps taken each day.

- ggplot(TotalSteps,aes(x=steps))+geom_histogram(fill="blue",binwidth=1000)+labs(title="Daily Steps",x="Steps",y="Frequency")

### Calculate and report the mean and median of the total number of steps taken per day

- TotalSteps[,.(MeanSteps=mean(steps,na.rm=TRUE),MedianSteps=median(steps,na.rm=TRUE))]

### What is the average daily activity pattern?

- Time series plot of the average number of steps taken.

- Make a time series plot (i.e. type="l") of the 5 minute interval (x-axis) and the average number of steps taken, average across all days (y-axis).

- IntervalDT <- activityDT[,c(lapply(.SD,mean,na.rm=TRUE)),.SDcols=c("steps"),by=.(interval)]

- ggplot(IntervalDT,aes(x=interval,y=steps))+geom_line(color="blue",size=1)+labs(title="Avg.Daily Steps",x="Interval",y="Avg. Steps per day")

### The 5-minute interval that, an average, contains the maximum number of steps

- IntervalDT[steps==max(steps),.(max_interval=interval)]

### Code to describe and show a strategy for imputing missing data

- Total number of missing values in the data set.

- sum(is.na(activityDT))

### Alternative Solution

- activityDT[is.na(steps),.N]

### Alternative Solution

- nrow(activityDT[is.na(steps),])

### Devise a strategy for filling in all of the missing values in the dataset. The strategy need not be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5 - minute interval etc

#### Filling in missing values with median of the data set

- activityDT[is.na(steps),"steps"]

- activityDT[,c(lapply(.SD,median,na.rm=TRUE)),.SDcols=c("steps)]

#### Create a new data set that is equal to the original data set but with the missing data filled in 

- data.table::fwrite(x=activityDT,file="data/tidyData.csv",quote=FALSE)

#### Histogram of the total number of steps taken each day after missing values are imputed

- Total number of steps taken per day

- TotalSteps <-activityDT[,c(lapply(.SD,sum)),.SDcols=c("steps"),by=.(date)]

#### Mean an median total number of steps taken per day

- TotalSteps[,.(MeanSteps=mean(steps),MedianSteps=median(steps))]

- ggplot(TotalSteps,aes(x=steps))+geom_histogram(fill="blue",binwidth=1000)+labs(title="Daily Steps",x="Steps",y="Frequency")

### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

- Determine if there are differences in activity patterns between weekdays and weekends?

- Create a new factor variable in the data set with two levels-"Weekday" and "Weekend" indicating whether a given date is a weekday or weekend day.

- Check date variable

- class(date)

#### Write date in POSIXct format

- activityDT[,date:=as.POSIXct(date,format="%Y-%m-%d)]

- activityDT[,"Day of Week":=weekdays(x=date)]

- activityDT[grepl(pattern="Monday|Tuesday|Wednesday|Thursday|Friday",x="Day of Week"),"weekday or weekend"]<-"weekday"

- activityDT[grepl(pattern="Saturday|Sunday",x="Day of Week"),"weekday or weekend"]<-"Weekend"

#### Convert weekday or weekend to a factor variable

- activityDTWDWD <-activityDT[,"weekday or weekend":=as.factor("weekday or weekend")]

- class(activityDTWDWE)

- levels(activityDTWDWE)

- head(activityDTWDWE,10)

### Panel plot (Time Series plot)

- activityDTWDWE[is.na(steps),"steps"]<-activityDTWDWE[,c(lapply(.SD,median,na.rm=TRUE)),.SDcols=c("steps")]

- IntervalDTWDWE <-activityDTWDWE[,c(lapply(.SD,mean,na.rm=TRUE)),.SDcols=c("steps"),by=.(interval,"weekday or weekend")]

- ggplot(IntervalDTWDWE,aes(x=interval,y=steps,color="weekday or weekend"))+geom_line()+labs(title="Avg. Daily Steps by Weektype",x="Interval",y="No. of Steps")+facet_wrap(~"weekday or weekend",ncol=1,nrow=2)

- All the R codes needed to reproduce the results (numbers,plots,etc) in the report.

### How to impute missing data

- Ways to approach missing data.

1. Ignore it - which means your software is choosing for you.

2. Imputation simply means replacing the missing values with an estimate, and then analyze the full data set as if the imputed values were actual observed values.

### Mean Imputation

- Simply calculate the mean of the observed values for that variable for all individuals who are non-missing.

- Advantages

1. Keeps the same mean.

2. Keeps the same sample size.

- Disadvantages

1. All single imputation methods underestimate standard errors.

2. They also underestimate p-values.

## Change working directory in RStudio

- Session.

- Set working directory.

- Choose directory.

- Alternative approach;

1. path <-getwd()

2. url <-"https://.....zip"

3. download.file(fileUrl,file.path(path,"dataFiles.zip"))

4. unzip(zipfile="dataFiles.zip")

- Alternative approach 2;

1. setwd("~/Desktop/datasciencecoursera/4_Exploratory_Data_Analysis/project/data")

## Some extra notes

### Sum a numeric variable with respect to factor variables

1. Use tapply

- tapply(NumVar,factor or categoricalVar,Sum)

- tapply(NumVar,catVar,mean)

2. Use lapply

- lapply(x,mean)

- lapply(x,quantile,probs=1:3/4)














