---
title: "Color in R Plots"
author: "David Asare Kumi"
date: "June 12, 2019"
output: html_document
---


##Part 1

###Plotting and color

- There are functions in R and in external packages that are very handy.

1. color 1 is Black.

2. color 2 is Red.

3. color 3 is Green.


- Default Image Plots in R from volcanoes.

1. heat.colors()

2. topo.colors()

- The grDevices package has two functions.

1. colorRamp.

2. colorRampPalette.

- The function colors() lists the names of colors you can use in any plotting function.


##colorRamp

There are 256 total numbers that you can specify for each color. colorRamp take a palette of colors and return a function that takes values between 0 and 1, indicating the extremes of the color palette (e.g see the "gray" function).

```{r,echo=TRUE,message=FALSE,warning=FALSE}
pal <-colorRamp(c("red","blue"))
pal(0)
pal(1)
pal(0.5)
pal(seq(0,1,len=10))

```

##colorRampPalette

colorRampPalette takes a palette of colors and return a function that takes integer arguments and returns a vector of colors interpolating the palette (like heat.colors or topo.colors).

```{r,echo=TRUE,message=FALSE,warning=FALSE}
pal <-colorRampPalette(c("red","yellow"))
pal(2) #hexadecimal

```

##There are 3 types of palettes

- Sequencial - for ordered data.

- Diverging - data that can deviate.

- Qualitative - data that are not ordered (factors,categorical)

##RColorBrewer and colorRampPalette

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(RColorBrewer)
cols <-brewer.pal(3,"BuGn")
cols

pal <-colorRampPalette(cols)

image(volcano,col=pal(20))

```

##The Smooth Scatter function

Another function that uses the color brewer palette is the smooth scatter function. This is useful for plotting (scatter) of a lot of different points.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
x <-rnorm(10000)
y <-rnorm(10000)
smoothScatter(x,y) #The default color is the blues

```

##Some other Plotting Notes

- The rgb function can be used to produce any color via red,green,blue proportions.

- Color transparency can be added via the alpha parameter to rgb.

- The colorspace package can be used for a different control over colors.

##Scatterplot with no transparency and Scatterplot with transparency

```{r,echo=TRUE,message=FALSE,warning=FALSE}
x <-rnorm(10000)
y <-rnorm(10000)
plot(x,y,pch=20)

#Scatterplot with transparency
plot(x,y,col=rgb(0,0,0,alpha=0.2),pch=20)

```
