---
title: "data.table"
author: "David Asare Kumi"
date: "June 13, 2019"
output: html_document
---


##What is data.table::as.data.table

- You can use data.table's fast and friendly file reader fread to load data directly as follows (data <-fread("data.csv")).

- fread accepts https and http URLs directly as well as operating system commands such as sed and awk output. See ?fread for example.

- data.table is an R package that provides an enhanced version of data frames, which are the standard data structure for storing data in base R. We already created a data.table using fread. We can also create one using data.table() function.

- You can also convert existing objects to a data.table using setDT() for (dataframes and lists()) and as.data.table() for other structures. See ?setDT() and ?as.data.table for more details.

##Subset rows in i

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
names(mpg)
class(mpg)
library(data.table)
mpgDT <-as.data.table(mpg)
class(mpgDT)
#get all cars with "4" as drv and class as compact
ans1 <-mpgDT[drv=="4" & class=="compact"]
ans1

#get all cars with "r" as drv and 1999 as year
ans11 <-mpgDT[drv=="r" & year==1999,] #in data.table you can ignore the comma
ans11

```

##Get the first five rows from your data

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
mpgDT <-as.data.table(mpg)
ans2 <-mpgDT[1:5]
ans2
ans22 <-mpgDT[1:5,]
ans22

```

##Sort data first by column hwy in ascending order, and then by year in descending order.

- We can use the R function order() to accomplish this task.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans3 <-mpgDT[order(hwy,-year)]#We can use "-" to sort in descending order.
head(ans3,10)

```

##Select columns in j

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
#Select hwy column but return it as a vector
ans4 <-mpgDT[,hwy]
head(ans4,10)
#Select hwy column but return it as a data.table instead
ans44 <-mpgDT[,list(hwy)]
head(ans44,10)

```

##Select several columns and return them as a data.table

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans5 <-mpgDT[,.(model,displ,year,cyl,drv,hwy)]
head(ans5,10)
#Alternatively
ans55 <-mpgDT[,list(model,displ,year,cyl,drv,hwy)]
head(ans55,10)

```

##Select both hwy and cyl columns and rename them to Highway and Cylinder

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans6 <-mpgDT[,.(Highway=hwy,Cylinder=cyl)]#the new name is to the left of the equal to sign
head(ans6,10)
#Alternatively
ans66 <-mpgDT[,list(Highway=hwy,Cylinder=cyl)]
head(ans66,10)

```

##Compute or do in j (column)

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
#how many cars have had total gallons > 27
ans7 <-mpgDT[,sum((hwy + cty)>27)]
ans7
#how many cars have had total gallons < 27
ans77 <-mpgDT[,sum((hwy + cty)< 27)]
ans77

```

##Subset in i and do in j

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
#calculate the average hwy and cty gallons for all drv=4 for the year 1999
ans8 <-mpgDT[drv=="4"& year==1999,.(ave_hwy=mean(hwy),ave_cty=mean(cty))]
ans8
#calculate the average hwy and cty gallons for all cyl=8 for the year 2008
ans88 <-mpgDT[cyl=="8"& year==2008,.(ave_hwy=mean(hwy),ave_cty=mean(cty))]
ans88

```

##Special Symbol .N

- It simply returns the number of rows in the subset.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans9 <-mpgDT[cyl=="8" & year==2008,.N]
ans9

```

##How can I refer to columns by names in j (like in a data.frame?)

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
#select hwy and cty columns in the data.frame way
ans10 <-mpgDT[,c("hwy","cty")]
head(ans10)

```

##Select columns named in a variable using the .. prefix

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
select_cols <-c("hwy","cty")
ans11 <-mpgDT[,..select_cols]
head(ans11,10)

#select columns named in a variable using with=FALSE
ans12 <-mpgDT[,select_cols,with=FALSE]
head(ans12)

```

##We can deselect columns using "-" or !

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans13 <-mpgDT[,!c("hwy","cty")]
head(ans13)
ans14 <-mpgDT[,-c("hwy","cty")]
head(ans14)

```

##Select and deselect range of columns

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
#select range of columns
ans15 <-mpgDT[,cyl:hwy]
head(ans15)
#deselect range of columns
ans16 <-mpgDT[,-(cyl:hwy)]
head(ans16)
ans17 <-mpgDT[,!(cyl:hwy)]
head(ans17)

```

##Aggregations

###Grouping using by

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans18 <-mpgDT[,.(.N),by=.(year)]
ans18
#alternatively
ans19 <-mpgDT[,.(.N),by="cyl"]
ans19
ans20 <-mpgDT[,.N,by=year]
ans20

```

##How can we calculate the number of cars for each year for manufacturer "audi"

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans21 <-mpgDT[manufacturer=="audi",.N,by=year]
ans21
##substitute year with cyl
ans22 <-mpgDT[manufacturer=="audi",.N,by=cyl]
ans22
#extension of conditions
ans23 <-mpgDT[manufacturer=="dodge",.N,by=.(year,cyl)]
ans23
#averages
ans24 <-mpgDT[manufacturer=="dodge",.(ave_hwy=mean(hwy),ave_cty=mean(cty)),by=.(year,cyl,drv)]
ans24

```

##sort by: keyby

- Actually keyby does a little more than just ordering. It also sets a key after ordering by setting an attribute called sorted.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans25 <-mpgDT[manufacturer=="dodge",.(ave_hwy=mean(hwy),ave_cty=mean(cty)),keyby=.(year,cyl,drv)]
ans25

```

##Chaining

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans26 <-mpgDT[manufacturer=="dodge",.N,by=.(hwy,cty)][order(hwy,-cty)]
head(ans26,10)

```

##Expressions in by

- Can by accept expressions as well or does it just take columns?

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans27 <-mpgDT[,.N,by=.(hwy>23,cty>16)]
ans27

```

##Special symbol .SD

- .SD stands for subset of data. It's by itself a data.table that holds the data for the current group defined using by.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans28 <-mpgDT[,print(.SD),by=drv]
ans28
#To compute on multiple columns, we can use the lapply() function
ans29 <-mpgDT[,lapply(.SD,mean)]
ans29

```

##Specify just the columns that we would like to compute the mean() on

- .SDcols accepts either column names or column indices. For e.g. .SDcols=c("arr_delay","dep_delay") ensures that .SD contains only these two columns for each group.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans30 <-mpgDT[manufacturer=="dodge",lapply(.SD,mean),by=.(cyl,drv,year),.SDcols=c("hwy","cty")]
ans30

```

##Subset .SD for each group

- How can we return the first 2 rows for each year?

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans31 <-mpgDT[,head(.SD,2),by=year]
head(ans31)

```

##How can we concatenate columns a and b for each group in cyl?

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
ans32 <-mpgDT[,.(val=c(hwy,cty)),by=year]
ans32
#return value as a list column
ans33 <-mpgDT[,.(val=list(c(hwy,cty))),by=cyl]
ans33

```

##rbind and cbind

```{r,echo=TRUE,message=FALSE,warning=FALSE}
sport <-c("Hockey","Baseball","Football")
league <-c("NHL","MLB","NFL")
trophy <-c("Stanley Cup","Commissioner's Trophy","Vince Lombardi Trophy")
trophies1 <-cbind(sport,league,trophy)
trophies1
trophies2 <-data.frame(sport=c("Basket Ball","Golf"),league=c("NBA","PGA"),trophy=c("Larry Obrian Championship Trophy","Wanamaker Trophy"),stringsAsFactors=FALSE)
#combine them into one data.frame with rbind
trophies <-rbind(trophies1,trophies2)
trophies

```

##Merge data.table


```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(ggplot2)
data(mpg)
library(data.table)
d1 <-mpgDT[,c("hwy","cty")]
head(d1)
d2 <-mpgDT[,c("displ","year")]
head(d2)
ans34 <-cbind(d1,d2)
ans34
#change data.table to data.frame
d1 <-as.data.frame(d1)
d2 <-as.data.frame(d2)
ans35 <-merge(d1,d2)
dim(ans35)
head(ans35)
#check the literature on merge. There is no common column name for the two sets of data

```