---
title: "My Fancy Presentation"
author: "David Asare Kumi"
date: "May 31, 2019"
output: ioslides_presentation
---


## My New Slide

- R markdown is built into Rstudio and allows you to create documents like HTML,PDF,and word documents from R. With R markdown you can embed R code into your documents.

### Why use R markdown?

- Turn work in R into more accessible formats.

- Incorporate R code and R plots into documents.

- R markdown documents are reproducible - the source code get re-run every time a document is generated, so if data changes or source code changes, the output in the document will change with it.

## Some r Code

```{r,comment="",echo=TRUE}
library(datasets)
data(mtcars)
head(mtcars)
dim(mtcars)

```

## Scatterplot

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
plot(mtcars$wt,mtcars$mpg,xlab="Weight",ylab="MPG",col="red",cex=2,
     pch=20,main="Scatterplot of MPG on WT")
fit<-lm(mtcars$mpg~mtcars$wt,data=mtcars)
abline(fit,lwd=2,lty=2,col="blue")

```

## Regression coefficients

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
summary(fit)$coef

```

## Diagnostic plots

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
par(mfrow=c(2,2))
plot(fit)

```

##Publishing and Sharing R markdown documents

- We publish in Rpubs.

- GitHub

- Dropbox

- GitHub Pages

- NeoCities
